---
title: "Purposes of regression analysis - Excercises "
output:
  html_document:
    code_foldin: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
```

We distinguish four types of modeling approaches, differentiated by the purpose of our analysis:

- Descriptive modeling
- Sample-to-population inferential modeling
- Explanatory modeling
- Predictive modeling

What type of modeling strategies best apply for the following questions? To see the correct answer click the `CODE` button.



1) What is the (causal) effect of raising the minimum wage on the unemployment rate?

```{r question-1}
"Explanatory"
```

2) How much revenue can we expect the upcoming advertisement campaign to generate?

```{r question-2}
"Predictive"
```

3) Are single mothers in the Czech Republic more likely to be at the risk of poverty than mothers with husbands?
  
```{r question-3}
"Sample-to-population inference"
```

4) Do regions in the Czech Republic with high unemployment rate also have higher children mortality?

```{r question-4}
"Descriptive"
```

5) Is there a relationship between education and pro-environmental attitudes in the population of Czech adults?

```{r question-5}
"Sample-to-population inference"
```