---
title: "Purposes of regression analysis - Excercises "
output:
  html_document:
    code_foldin: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
```

We know of four types of models, based on the the purpose of our analysis:

- Descriptive models
- Inferential models
- Causal/explanative models
- Predictive models

What type of model we would use to answer the following questions? To see the correct answer click the `CODE` button.



1) What is the effect of raising the minimum wage on the unemployment rate?

```{r question-1}
"Causal/explanative model"
```

2) How much revenue can we expect the upcoming advertisement campaign to generate?

```{r question-2}
"Predictive model"
```

3) Are women on parental leave in the Czech Republic more likely to be at the risk of poverty than men?
  
```{r question-3}
"Inferential model"
```

4) Do regions in the Czech Republic with high unemployment rate also have high children mortality?

```{r question-4}
"Descriptive model"
```

5) Is there a relationship between education and pro-environmental attitudes in the population of Czech adults?

```{r question-5}
"Inferential model"
```