---
title: "Simple linear regression - Excercises "
output:
  html_document:
    code_foldin: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
```

```{r data}
countries = read.csv("data/countries.csv")
```

1) Using the `countries` data, create a regression model predicting the percentage of people at risk of poverty (`poverty_risk`) by using the democratic index (`dem_index`).

```{r}
mod1 = lm(poverty_risk ~ dem_index, data = countries)
```

2) Interpret regression coefficients of the model.

```{r}
summary(mod1)

"Intercept - the expected proportion of people at the risk of poverty in a country with democratic index of zero. In such country, we would expect to see 63% (0.63) of people at the risk of poverty"

"dem_index - the difference in the expected proportion of people at the risk of povery when the democratic index raises by one. According to our model, countries with 1 additional point on the democratic index have on average 5 percentage points smaller proportion of people at the risk of poverty"
```

3) Does the intercept have a meaningful interpretation in our model?

```{r}
"No, as no country have a democratic index of zero"
```


4) Center the democratic index variable and fit the model again. How does the intepretation of the coefficients changes?

```{r}
countries$dem_index_c = countries$dem_index - mean(countries$dem_index, na.rm = TRUE)

mod2 = lm(poverty_risk ~ dem_index_c, data = countries)
summary(mod2)

"The intercept now represents the proportion of people at the risk of poverty in a country with average democratic index. The interpretation of the slope coefficient does not change."
```