---
title: "Simple linear regression - Excercises "
output:
  html_document:
    code_foldin: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
```

```{r data}

library(tidyverse)
countries = read.csv("data/countries.csv")
```

1) Using the `countries` data, create a regression model predicting the percentage of people at risk of poverty (`poverty_risk`) by using the democracy index (`dem_index`).

```{r}

countries %>% ggplot(aes(x=dem_index, y=poverty_risk))+geom_point()+geom_smooth(method = "lm") # remember to visualize before modeling 

mod1 = lm(poverty_risk ~ dem_index, data = countries)
```

2) Interpret regression coefficients of the model.

```{r}
summary(mod1)

"Intercept - the expected proportion of people at the risk of poverty in a country with the democracy index of zero. In such country, we would expect to see 63% (0.63) of people at the risk of poverty."

"dem_index - the difference in the expected proportion of people at the risk of povery when the democracy index raises by one. According to our model, countries with 1 additional point on the democracy index have on average 5 percentage points smaller proportion of people at the risk of poverty"
```

3) Is the interpretation of the intercept in our model (as stated in previous solution) meaningful?

```{r}
"No, it is only technical. On inspection of the data, you will notice there are no countries in the sample even close to the democracy index of 0. So the interpetation is, in fact, a wild extrapolation to an area for which we have no data."
```


4) Center the democracy index variable and fit the model again. How does the interpretation of the coefficients changes?

```{r}
countries$dem_index_c = countries$dem_index - mean(countries$dem_index, na.rm = TRUE)

mod2 = lm(poverty_risk ~ dem_index_c, data = countries)
summary(mod2)

"The intercept now represents the proportion of people at the risk of poverty in a country with average democracy index (average within our sample, of course). This is much more useful. The interpretation of the slope coefficient does not change."
```

5) Use the categorized democracy index instead, i.e., the variable di_cat as predictor. What do you need to do to include a categorical predictor in linear regression analysis? Interpret your model.


```{r}
library(tidyverse) # for pipes, count function and ggplot

str(countries$di_cat) # check that the variable is not coded as numeric
countries %>% count(di_cat) # out of curiosity, we look at a basic frequency table

mod3 = lm(poverty_risk ~ di_cat, data = countries)
summary(mod3)

"We did not have to do anything special before running the model, just check that that the di_cat variable is not coded as numeric. Since it is not, R automatically transforms it into factor (if it is not a factor yet) and then to dummies. It automatically includes only k-1 of the dummies in the model. We see, looking at the intercepts, that full democracy have 6 percentage points less people living in poverty, on average, compared to Flawed democracies. Hybrid democracies, on the other hand, have 16 percentage points higher povery rate the Flawed democracies, on average. But what is the poverty rate of Flawed democracies to begin with? It is the intercept, i.e. 25.5 %."
```


