---
title: "Multiple linear regression"
output: html_document
csl: apa.csl
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

```{r data-and-packages, echo=FALSE}
library(tidyverse)
library(patchwork)
library(here)


countries <- read.csv(here("data", "countries.csv"))
```

The goals for this lecture are (1) understanding the concept of multiple
linear regression with emphasis on interpretation of its coefficients,
and (2) building and interpreting multiple regression models in R. 

# Introduction to multiple linear regression

Multiple linear regression (MLR) is defined as having multiple predictor terms (more than one). Remember the general formula for simple linear
regression:

$$
Y = \alpha + \beta*X + \epsilon,
$$
where $\alpha + \beta*X$ represents the line and $\epsilon$ represent the deviation from it. 


The formula for MLR is its fairly straightforward extension. Notice that
intercept in MLR is usually called $\beta_0$ rather than $\alpha$. The
other regression coefficients are numbered betas. The numbered x-es stand for predictor terms.

$$
Y = \beta_0 + \beta_1*x_1 + + \beta_2*x_2 + ... + \epsilon
$$

Appreciate that having more predictor terms does not necessarily imply using
more predictor variables. For example, the blue line in the following
figure represents linear model estimated using three predictor terms:
democratic index, democratic index to the power of two (quadratic term),
and democratic index to the power of three (cubic term). While the model
only uses one predictor variable, there are three predictor terms which
makes the model a case of MLR (the interpretation framework of MLR explained in this lecture applies). For this model, four betas would be
estimated (one of them the intercept).

```{r multiple1, message=FALSE, warning=FALSE, fig.width=5, fig.align="center", echo=FALSE}

countries %>%
  ggplot(aes(x= dem_index, y = life_exp)) +
  geom_point()+
  geom_smooth(method = "lm", se = FALSE, formula = y ~ poly(x,3)) +
  labs(x = "predictor variabel on x axis\nhere: democratic index 1-10",
       y = "response variable on y axis\nhere: life expectancy",
       caption = "life expectancy = democratic_index + democratic_index^2 + democratic_index^3")
```

One of the common misconceptions about linear regression is that
it can only estimate a straight line. That is only true for simple
linear regression. From the picture above, it should be clear that in
MLR, it is possible to "bend" the line. And not only that. Throughout the course we will see how flexible MLR can be.

Another example of a MLR model is visualized on the figure below. It
uses two predictor terms (in this case, they are two distinct
independent variables): again, the democratic index, and, in addition, a
binary variable whether the country is post-soviet or not. Due to the
fact that one of the independent variables is binary, it is possible to
visualize the whole model in one two-dimensional figure as two lines.
However, more complex MLR models cannot be meaningfully visualized
on top of the raw data.

```{r multiple2, message=FALSE, warning=FALSE, fig.width=6, fig.align="center", echo=FALSE}

countries %>%
  ggplot(aes(x= dem_index, y = life_exp, color = postsoviet)) +
  geom_point()+
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x) +
  labs(x = "predictor variabel on x axis\nhere: democratic index 1-10",
       y = "response variable on y axis\nhere: life expectancy",
       caption = "life expectancy = democratic index + postsoviet")

```

# Interpreting coefficients in MLR

First, let's run a model where life expectancy is explained with
democratic index and percentage of people with university degree in
15-64 age category (that is how the variable is defined so we just use what we have).

```{r model2, include=TRUE}

fit1 <- lm(life_exp ~ dem_index + uni_prc, data = countries)
summary (fit1)
```

The output above should look familiar, it differs from that of simple linear regression (previous lecture) only by having an extra row in the coefficient table. The beta coefficients are in the column called "Estimate".
Specifically, the coefficient for democratic index is 1.786 (recall that the coefficinet was 1.62 when democratic index was the only predictor), and the
coefficient for university education is -1.460.

Interpreting coefficients in MLR can be surprisingly tricky as they are,
"in part, contingent on the other variables in the model" [@gelman2020,
pp. 131]. This means that they will change when other variables are
added to the model or dropped from the model.

We will cite @gelman2020 on how best to interpret them: "The coefficient
$\beta_k$ is the average or expected difference in outcome $y_k$,
comparing two people who differ by one unit in the predictor $x_k$ while
being equal in all the other predictors. This is sometimes stated in
shorthand as comparing two people (or, more generally, two observational
units) that differ in $x_k$ with all the other predictors held constant"
(p. 131). We also call this conditional effect (as in "conditional on the assumption that all other predictors are held constant").

For the example above, we could say that if one country has democratic
index higher by one point than another country, and they both have the
same percentage of university educated people, the first country will
have expected life expectancy higher by 1.8 years (we encourage to reasonably round the beta of 1.786 for interpretation purposes to avoid making impression of unjustified precision).

Since the coefficient for percentage of university educated people has
very broad confidence interval (and very high p-value), we could also say that we find no
evidence in the data, that two countries with the same level of democratic
index should be expected to differ in life expectancy when they have different proportion of university educated people. That is if we believe
the sample of countries justifies some generalization to some concrete
or at least abstract population of countries. Someone could be tempted
to make a purely descriptive interpretation with no ambition to make
sample-to-population inference. He or she could try to say something
along the lines that for our particular sample of countries and data
from given years, we find that for two countries with the same level
democratic index the life expectancy goes down by 1.46 years, on average,
as the proportion of university educated increases from 0 to 100% (the
variable is measured on scale from 0 to 1). This seems to us a
meaningless statement and we would discourage from it, even if
technically true. The really large standard errors should warn us
against making interpretation even in purely descriptive situations.
Saying that controlling for democratic index, there does not seem to be
any linear association between life expectancy and proportion of university
educated in our sample of countries seems much more sensible.\

Within descriptive modeling strategies, we use MLR to see the effects of
individual variables net of the effects of all the other predictors in
the model. Within predictive modeling, we use MLR to improve our
predictions over simple linear model (predictions based on just one
predictor term often tend to be weak in social sciences). Within
explanatory modeling, we use MLR for adjusting for background variables according to our theory,
hence discovering potentially spurious relationships and isolating potentially causal effects.

## Difference-based vs. change-based interpretations

There are two conceptually slightly different interpretations of MLR
coefficients. The difference-based interpretation is well described as
"how the outcome variable differs, on average, when comparing two groups
of items that differ by 1 in the relevant predictor while being
identical in all the other predictors" [@gelman2020, pp. 134]. In contrast, the
change-based interpretation is well described by saying that "the
coefficient is the expected change in y caused by adding 1 to the
relevant predictor, while leaving all the other predictors in the model
unchanged" [@gelman2020, pp. 134].

In others words, the difference-based interpretation uses the idea of
difference between individuals, whereas the change-based interpretation
uses the idea of change within individual. To be on the save side, we
recommend interpreting regression coefficients as comparisons between
units, not (potential) changes within units, unless one specifically
claims causality. The difference-based interpretation is simply more
general, hence carries less risk of being used in a misleading way. (By misleading we mean invoking causality where no evidence for a causal relationship is shown.)

Note that [@gelman2020] use the term "predictive interpretation" instead
of difference-based interpretation and "counterfactual" interpretation
instead of change-based interpretation.

## Comparing beta coefficients of different predictors

It can be sometimes useful to compare coefficients of different
independent variables in the model. As we saw in the model above, it
cannot be done in a straightforward way. While values of 1.786 and 1.460
are not very far from each other, it makes no sense to compare them. The
first shows change when two countries differ in democratic index
(measured on the scale from 0 to 10) by one (a realistic situation),
whereas the latter shows change when one country has no university
educated people and the other has only university educated people (a
very unrealistic situation). Indeed, the coefficient value depends on
the unit we use for measuring the independent variable.

To deal with this issues of no comparability, we can use so called
standardized betas. Those can be used to determine relative weight or strength of
independent variables as they show the effect of an increase in X by one
standard deviation on Y, also measured in standard deviations. The way
to calculate standardized betas is to first standardize (compute
z-scores) all variables used in the model and then run a regression on
the z-scores.

A way to do this using the scale function directly in the regression call is shown below.

```{r model, include=TRUE}

fit2 <- lm(scale(life_exp) ~ scale(dem_index) + scale(uni_prc), data = countries)
summary (fit2)
```

Now we can compare the relative strengths of our predictors. The
democratic index (with standardized beta of 0.82) is a way stronger
predictor of life expectancy than the proportion of university educated
(0.04 in absolute value as the sign is irrelevant for this comparison).
Notice that the values of t-tests for the significance of coefficients
did not change. It should come as no surprise, we did not substantively
change the data by standardizing it. Also remember that we are comparing the net effect of the predictors. The comparison could look quite different when standardized betas from two simple linear regression models were compared.  

@gelman2008 suggests to use slightly different transformation if we also
want include binary predictors (dummy variables) in the model. He shows
that standardizing by subtracting the mean and dividing by 2 standard
deviations (rather than just 1) enables to directly compare this kind of
standardized betas with coefficients for non-standardized binaries.


# References
