---
title: "Multiple linear regression"
output: html_document
csl: apa.csl
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

```{r data-and-packages}
library(tidyverse)
library(patchwork)
library(here)


countries <- read.csv(here("data", "countries.csv"))
```

The goals for this lecture are (1) understanding the concept of multiple
linear regression with emphasis on interpretation of its coefficients,
and (2) building and interpreting multiple regression models in R. We
will also discuss coding categorical predictors as dummy variables and
using interaction.

# Introduction to multiple linear regression

Multiple linear regression (MLR) is defined by having more than one
predictor term. Remember the general formula for simple linear
regression:

$$
y = \alpha + \beta*x + \epsilon_i
$$

The formula for MLR is its fairly straightforward extension. Notice that
intercept in MLR is usually called $\beta_0$ rather than $\alpha$. The
other regression coefficients are numbered betas. The numbered x-es
stand for predictor terms.

$$
y = \beta_0 + \beta_1*x_1 + + \beta_2*x_2 + ... + \epsilon_i
$$

Note that having more predictor terms does not necessarily imply using
more predictor variables. For example, the blue line in the following
figure represents linear model estimated using three predictor terms:
democratic index, democratic index to the power of two (quadratic term),
and democratic index to the power of three (cubic term). While the model
only uses one predictor variable, there are three predictor terms which
makes the model a case of MLR. For this model, four betas would be
estimated (one of them the intercept).

```{r multiple1, message=FALSE, warning=FALSE, fig.width=5, fig.align="center", echo=FALSE}

countries %>%
  ggplot(aes(x= dem_index, y = life_exp)) +
  geom_point()+
  geom_smooth(method = "lm", se = FALSE, formula = y ~ poly(x,3)) +
  labs(x = "predictor variabel on x axis\nhere: democratic index 1-10",
       y = "response variable on y axis\nhere: life expectancy",
       caption = "life expectancy = democratic_index + democratic_index^2 + democratic_index^3")
```

One of the (many?) common misconceptions about linear regression is that
it can only estimate a straight line. That is only true for simple
linear regression. From the picture above, it should be clear that in
MLR, it is possible to "bend" the line. And not only that.

Another example of a MLR model is visualized on the figure below. It
uses two predictor terms (in this case, they are two distinct
independent variables): again, the democratic index, and, in addition, a
binary variable whether the country is post-soviet or not. Due to the
fact that one of the independent variables is binary, it is possible to
visualize the whole model in one two-dimensional figure as two lines.
However, more complicated MLR models cannot be meaningfully visualized
on top of the raw data like this. What we can do is visualize their
results (apart from displaying them as numbers in a table). We will
discuss later how it is done, but we should first make clear how the
beta coefficients are interpreted in MLR.

```{r multiple2, message=FALSE, warning=FALSE, fig.width=6, fig.align="center", echo=FALSE}

countries %>%
  ggplot(aes(x= dem_index, y = life_exp, color = postsoviet)) +
  geom_point()+
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x) +
  labs(x = "predictor variabel on x axis\nhere: democratic index 1-10",
       y = "response variable on y axis\nhere: life expectancy",
       caption = "life expectancy = democratic index + postsoviet")

```

# Interpreting coefficients in MLR

First, let's run a model where life expectancy is explained with
democratic index and percentage of people with university education in
15-64 age category (that is how the variable is defined).

```{r model2, include=TRUE}

fit1 <- lm(life_exp ~ dem_index + uni_prc, data = countries)
summary (fit1)
```

The output in the frame with white background above is the raw summary
output which appears in R console as response to running the two lines
of code. The beta coefficients are in the column called "Estimate".
Specifically, the coefficient for democratic index is 1.786, and the
coefficient for university education is -1.460.

Interpreting coefficients in MLR can be surprisingly tricky as they are,
"in part, contingent on the other variables in the model" [@gelman2020,
pp. 131]. This means that they will change when other variables are
added to the model or dropped from the model.

We will cite @gelman2020 on how best to interpret them: "The coefficient
$\beta_k$ is the average or expected difference in outcome $y_k$,
comparing two people who differ by one unit in the predictor $x_k$ while
being equal in all the other predictors. This is sometimes stated in
shorthand as comparing two people (or, more generally, two observational
units) that differ in $x_k$ with all the other predictors held constant"
(p. 131). We also call this conditional effect.

For the example above, we could say that if one country has democratic
index higher by one point than another country, and they both have the
same percentage of university educated people, the first country will
have life expectancy higher by 1.8 years (i.e., rounding 1.786), on
average.

Since the coefficient for percentage of university educated people has
very broad confidence interval, we could also say that we find no
evidence in the data, that two countries with the same level democratic
index should be expected to differ in life expectancy conditional on
their proportion of university educated people. That is if we believe
the sample of countries justifies some generalization to some concrete
or at least abstract population of countries. Someone could be tempted
to make a purely descriptive interpretation with no ambition to make
sample-to-population inference. He or she could try to say something
along the lines that for our particular sample of countries and data
from given years, we find that for two countries with the same level
democratic index the life expectancy goes down by 1.5 years, on average,
as the proportion of university educated increases from 0 to 100% (the
variable is measured on scale from 0 to 1). This seems to us a
meaningless statement and we would discourage from it, even if
technically true. The really large standard errors should warn us
against making interpretation even in purely descriptive situations.
Saying that controlling for democratic index, there does not seem to be
any association between life expectancy and proportion of university
educated in our sample of countries seems much more sensible.\
Within descriptive modeling strategies, we use MLR to see the effects of
individual variables net of the effects of all the other variables in
the model. Within predictive modeling, we use MLR to improve our
predictions over simple linear model (predictions based on just one
predictor term often tend to be weak in social sciences). Within
explanatory modeling, we use MLR for adjusting for background variables,
hence discovering potentially spurious relationships.

## Difference-based vs. change-based interpretations

There two conceptually slightly different interpretations of MLR
coefficients. The difference-based interpretation is well described as
"how the outcome variable differs, on average, when comparing two groups
of items that differ by 1 in the relevant predictor while being
identical in all the other predictors" [@gelman2020, pp. 134]. The
change-based interpretation is well described by saying that "the
coefficient is the expected change in y caused by adding 1 to the
relevant predictor, while leaving all the other predictors in the model
unchanged" [@gelman2020, pp. 134].

In others words, the difference-based interpretation uses the idea of
difference between individuals, whereas the change-based interpretation
uses the idea of change within individual. To be on the save side, we
recommend interpreting regression coefficients as comparisons between
units, not (potential) changes within units, unless one specifically
claims causality. The difference-based interpretation is simply more
general, hence carries less risk of being used in a misleading way.

Note that [@gelman2020] use the term "predictive interpretation" instead
of difference-based interpretation and "counterfactual" interpretation
instead of change-based interpretation.

## Comparing beta coefficients of different predictors

It can be sometimes useful to compare coefficients of different
independent variables in the model. As we saw in the model above, it
cannot be done in a straightforward way. While values of 1.786 and 1.460
are not very far from each other, it makes no sense to compare them. The
first shows change when two countries differ in democratic index
(measured on the scale from 0 to 10) by one (a realistic situation),
whereas the latter shows change when one country has no university
educated people and the other has only university educated people (a
very unrealistic situation). Indeed, the coefficient value depends on
the unit we use for measuring the independent variable.

To deal with this issues of no comparability, we can use so called
standardized betas. Those can be used to determine relative weight of
independent variables as they show the effect of an increase in X by one
standard deviation on Y, also measured in standard deviations. The way
to calculate standardized betas is to first standardize (compute
z-scores) all variables used in the model and then run a regression on
the z-scores.

A way to do this using the scale function is shown in the code below.

```{r model, include=TRUE}

fit2 <- lm(scale(life_exp) ~ scale(dem_index) + scale(uni_prc), data = countries)
summary (fit2)
```

Now we can compare the relative strengths of our predictors. The
democratic index (with standardized beta of 0.82) is a way stronger
predictor of life expectancy than the proportion of university educated
(0.04 in absolute value as the sign is irrelevant for this comparison).
Notice that the values of t-tests for the significance of coefficients
did not change. It should come as no surprise, we did not substantively
change the data by standardizing it.

@gelman2008 suggests to use slightly different transformation if we also
want include binary predictors (dummy variables) in the model. He shows
that standardizing by subtracting the mean and dividing by 2 standard
deviations (rather than just 1) enables to directly compare this kind of
standardized betas with coefficients for non-standardized binaries.

# Main effects and interactions

All the beta coefficients in the examples above were so called main
effects. We use this term to distinguish them from interactions which we
will introduce in this section. First about the main effects. There are
two types of main effects. First, in simple linear regression (with one
predictor term), the main effect indicates the bivariate linear
association between independent and dependent variable. Second, in MLR,
the main effect is the conditional effect described above (i.e., the
effect of independent variable on a dependent variable conditional on
all other variables in the model being held constant).

Using only the main effects in MLR relies on the assumption that the
effect of a given independent variable on the dependent variable is the
same across all levels of other independent variables in the model, or
more precisely, that rather than looking at different effects of our
independent variable of interest for different levels of other
independent variables, it is an acceptable simplification to take the
average effect.

This is easier explained with an example. Imagine you model the effect
of years of education on earnings and you also enter gender into the
model (we assume it only takes two values, "man" or "woman". The
equation would like like this: \$ earnings = \beta\_0 +
\beta\_1\*yearsOfEducation + \beta\_2\*gender\$. The coefficient
$\beta_1$ shows the conditional effect of years of education on income
for some hypothetical average gender. This can be an acceptable
simplification if the effect is similar for both men and women. But
imagine there is strong positive effect of years of education on income
for one gender and similarly strong negative effect for the other
gender. These effects will cancel out (provided there is about the same
number of men an women in the sample) and the $\beta_1$ will be close to
zero. This would be a problematic simplification of the reality - we
would lose a potentially important piece of information. To prevent
this, we can use interactions.

An interaction enables to show how the effect of one independent
variable on the dependent variable varies across levels of another
independent variable. In the example above, the equation would look like
this:
$income = \beta_0 + \beta_1*yearsOfEducation + \beta_2*gender + \beta_3*yearsOfEducation*gender$.
The interpretation of coefficients is prone to errors, caution is
advised. The intercept is the value of income for the reference category
of gender (say man as R would code them as 0s because, alphabetically,
they come before women) with 0 years of education. The coefficient for
gender would show the difference for between the predicted income for
men in women with no years of education. The coefficient for years of
education can be thought of as the comparison of average income for men
who differ by one year of education. Finally, the interaction
coefficient represents the difference in slope for years of education
between men and women.

[TBD - TABLE AND FIGURE TO ILUSTRATE THIS, EXAMPLE OF EQUATION FOR
SELECTED INDIVIDUALS]

## Model with interaction VS. two separate models

If we are interested in an interaction between two continuous variables, we have no other option than running a model with interaction. However, when we are interested in an interation between one continuous and one binary variable, we could also be tempted to run two separate models, one for men and one for women. This would not be a horrible solution. The estimated effected sizes would be the same. However, we would not be able to calculate the overall fit of the model (we will talk about model fit in one of the next sessions), and we would not have the estimate of standard error of the difference of slopes between men and women. So unlike in the model with interaction, we would not be able to say if the didference in slope between the two genders is statistically significant. 


## Centering variables in models with interaction

We saw above that some of the interpretations of main effects can be difficult or not very useful when models contain interactions. Indeed, they are estimated with the other independent variables fixed at 0. And the 0 can be very unrealistic such as in case of years of education above. 

However, we can use centering for better interpretability of main effects. After doing so, the main effect corresponds to a predictive difference with the other input fixed at its average value. Intercept in the example above would be the average income for a person with mean number of years of education and of "average gender". Coefficient for years of education would be the average slope for all levels of gender. The interaction would keep the ame interpretation as above (difference of slope between genders).         

[TBD - TABLE AND FIGURE TO ILUSTRATE THIS]

## Interaction as "it depends effects"

Jim Frost offers in his [blog](https://cutt.ly/HkXcMY4) a helpful way to
think about interactions as of "it depends effects". If the answer to
the question 'What is the effect of X on Y?' is 'It depends', we are
dealing with interactions. For example, do you like your food more with
ketchup added? It depends, yes, if the food is french fries, no, if the
food is ice-cream sundae.

## Interactions require big samples

Adding interactions sometimes makes good sense theoretically, and
sometimes it also helps achieve good model specification with good fit
to the data. However, estimating interactions requires much bigger
samples to identify the same effect size as for main effect. And since
we can mostly assume that the effect size of interactions will be
smaller than that of main effects, we will mostly require MUCH bigger
samples. In his
[blog](https://statmodeling.stat.columbia.edu/2018/03/15/need-16-times-sample-size-estimate-interaction-estimate-main-effect/),
Andrew Gelman shows that we will need 16 times the sample size to
estimate an interaction which is half the size of the main effect if we
want to have the same statistical power. This does not mean that
interactions are beyond reach, we can still meaningfully estimate them
when their effect size is large enough. But when their effect size is
only small, they can be out of reach in common sociological samples.
When we look for interactions in explorative analysis, we "typically
look for them is with predictors that have large coefficients when not
interacted. For a familiar example, smoking is strongly associated
with cancer. In epidemiological studies of other carcinogens, it is crucial to adjust for smoking both as
an uninteracted predictor and as an interaction, because the strength of association between other risk
factors and cancer can depend on whether the individual is a smoker." [@gelman2020, pp. 136].


# References
