---
title: "Visualizing regression models"
subtitle: "Jaromír Mazák & Aleš Vomáčka"
author: "Dpt. of Sociology, Faculty of Arts, Charles University"
date: "Last updated in March 2022"
output:
  ioslides_presentation:
    widescreen: true
    incremental: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(here)
library(ggeffects)
library(broom)

countries <- read.csv(here("data", "countries.csv"))
```

## Goals for today

- Learning how to produce point-range regression plots
- Understanding marginal effects 
- Plotting marginal effects

# Point-range plots

## Point-range plots

- Straightforward visualizations of the regression coefficients tables

```{r forest-example, fig.height=3.5}
m1 = lm(life_exp ~ hdi + dem_index + poverty_risk, data = countries)

tidy(m1) %>% 
  ggplot(aes(x = estimate, 
             xmin = estimate - 1.96*std.error, 
             xmax = estimate + 1.96*std.error,
             y = term)) +
  geom_pointrange()
```

## Standardization

- Visual display invites comparison, so better standardize

```{r forest-example-z, fig.height=3.5}

m2 = lm(scale(life_exp) ~ scale(hdi) + scale(dem_index) + scale(poverty_risk), data = countries)

tidy(m2) %>% 
  ggplot(aes(x = estimate, 
             xmin = estimate - 1.96*std.error, 
             xmax = estimate + 1.96*std.error,
             y = fct_reorder(term, estimate))) +
  geom_pointrange()
```

# Marginal effect plots


## Some models are easy to interpret...

- Relationship between life expectancy and democratic index, assuming linear relationship

```{r linear-simple, echo=TRUE}
mod1 = lm(life_exp ~ dem_index, data = countries)
coef(mod1)
```

- A country with additional point of democratic index has, on average, 1.6 years higher life expectancy

## ... others not so much

- Relationship between life expectancy and democratic index, but the relationship can be parabolic and also can differ based on whether the country has a post soviet history or not.

```{r linear-not-simple, echo=TRUE}

m2 = lm(life_exp ~ poly(dem_index, 2) * postsoviet,
          data = countries[!is.na(countries$dem_index),])
coef(m2)
```

- Help, how can I interpret this?

## Complex models are hard to interpret 

- Once variables are transformed, there are interactions, quadratic terms, splines, etc., interpreting relationships between them using coefficients tables becomes difficult
- Solution: visual interpretation using marginal effects

## Marginal effects are here to save the day

- Marginal effect = expected value of the dependent variable for given values of selected independent variable while other independent variables are held constant (preferably at some reasonable value such as mean)

```{r marg-eff-table, echo=TRUE}
ggpredict(m2, terms = c("dem_index [4,6,8]", "postsoviet [yes]"))
```

## Marginal effects vizualized

- The plot contains the exact same information as the table of regression coefficients, but is actually interpretable

```{r marg-eff-plot}
plot(ggpredict(m2, terms = c("dem_index [all]", "postsoviet")))
```


## Marginal effects

- We can also plot just one variable one, while holding the others constant
- The following plot shows the difference in expected life expectancy for countries with **average** democratic index

```{r marg-eff-plot-solo, fig.height=3}
plot(ggpredict(m2, terms = c("postsoviet")))
```

- Question: How is this plot different from displaying simply group means, i.e. mean life expectancy for the postsoviet vs. non-postsoviet?
  
## `ggeffects` package

- `ggeffects` package computes marginal effects and returns `ggplot2` friendly data frames

- The `ggeffects` offers three basic functions:

  - `ggpredict` - based on the base `predict()` function, works on almost any model
  - `ggeffects` - based on the `effects` package, similar to ggpredict, but handles categorical variables differently
  - `ggemeans` - based on the `emeans` package, mostly for contrasts


## `ggeffects` package is broadly applicable

- The package works for a variety of models not covered in this course
  - generalized linear models
  - (generalized) mixed models
  - bayesian models
  - survival models
  - and many more

And that is pretty cool!
  

