---
title: "Linearity: The rehearsal"
subtitle: "Jaromír Mazák & Aleš Vomáčka"
author: "Dpt. of Sociology, Faculty of Arts, Charles University"
date: "Last updated in April 2021"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r data-and-packages, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(splines)
library(lspline)
library(ggeffects)
library(patchwork)

vote = read.csv("../data/parl_vote_2017.csv")
un = read.table("../data/UnitedNations.txt")
```

## What linearity?

- People talk about linearity into two ways:
  - linear relationships between variables
  - linear forms of regression models
  
## Linear relationships

```{r}
d1 = tibble(x = seq(1,100,1),
               y = 2*x)

d2 = tibble(x = seq(1,100,1),
            y = x^2)

g1 = qplot(x = x,y = y, data = d1, geom = "line")
g2 = qplot(x = x,y = y, data = d2, geom = "line")

g1 + g2
```

## Linear forms

- Linear forms

$vote = \beta_0 + \beta_1 *age$

$vote = \beta_0 + \beta_1 *age + \beta_2*age^2$

$log(income) = \beta_0 + \beta_1 *log(age) + \beta_2*log(education)$

- Nonlinear forms

$price = \beta_0 + \frac{1}{\beta_1} * demand$

$income = \beta_0 + \beta_1*age^{\beta_2*education}$

## Formal linearity

- Mathematically, linear regression assumes that the linear form can capture the relationships between variables
- e.g.:

$log(income) = \beta_0 + \beta_1 *log(age) + \beta_2*log(education)$

can capture the relationship between income, age and education (whether the relationship *between the raw variables* is linear or not)


- However just because your linear model can capture **a** nonlinear relationship, it won't necessarily capture the correct relationship
- The residual plot basically tells you if the relationship between variables differs from the relationships assumed by your model

